---
title: "Workshop Genome Wide CNVs in the UKB"
author: "Simone Montalbano"
output:
  html_document:
    toc: true
    theme: cerulea
---

```{r setup, include = F}
knitr::opts_chunk$set(warning = FALSE, echo = TRUE, eval = TRUE,
                      include = TRUE, results = "markup", cache = T)
```

```{r remove cache, eval = F, include = F}
# not eval, run manually to remove the cached files
unlink('./analysis_cache', recursive = TRUE)
unlink('./analysis_files', recursive = TRUE)
```

```{r compile html, eval = F, include = F}
# not eval, run manually to compile into an HTML
rmarkdown::render("analysis.Rmd")
```


```{r QCtreeCNV, eval = F, include = F}
# install QCtreeCNV package from GitHub if needed
devtools::install_github("sinomem/QCtreeCNV")
```

# Data and libraries

## Libs

Load necessary libraries and CNValidatron package.

```{r libs, cache = F}
library(data.table)
library(ggplot2)
```

```{r, eval = T}
# load the package, it will require additional installations
setwd('../CNValidatron_fl')
devtools::load_all()
setwd('../NSHG_CNVs_workshop_2024')
```

```{r, eval = F}
# alternatively, just load the functions we use
# way less safe and far from optimal but should work
list.files('../CNValidatron_fl/R')
for (f in list.files('../CNValidatron_fl/R'))
  source(paste0('../CNValidatron_fl/R/', f))
```


## CNV table

Load simulated CNVs data.table.

```{r load CNVs}
cnvs <- fread('cnvs.tsv')
```


## Phenotypes

Load simulated pheno file.

```{r load pheno}

```



# How to analyse Genome Wide CNVs

As you might have imagined from the IGV session, CNVs can be a tricky class
of variants to analyse. Their size spans from tens kilobases to few megabases
and their boundaries can vary a lot even among "similar" CNVs.

There are two main approaches to solve this problem. Use some kind of bins of
the genome, such as regular windows or genes, or compute CNV regions (CNVRs).
In this section we will cover both approaches.


## Binning the genome


```{r bin cnvs, eval = F}
dt <- binned_cnvs(cnvs, format = 'count', bins = CNValidatron:::binned_genome(bin_size = 250000))
saveRDS(dt, 'binned_cnvs_250kbp.rds')
dt <- binned_cnvs(cnvs, format = 'count', bins = CNValidatron:::binned_genome(bin_size = 100000))
saveRDS(dt, 'binned_cnvs_100kbp.rds')
```

```{r binned BED, eval = F}
dt <- CNValidatron:::binned_genome(bin_size = 250000)[]
dt[, end := end + 2]
fwrite(dt[, .(chr, start, end, ix)], 'bed_files/bins_250.bed',
       sep ='\t', col.names = F, scipen = 1000)

dt <- CNValidatron:::binned_genome(bin_size = 100000)[]
dt[, end := end + 2]
fwrite(dt[, .(chr, start, end, ix)], 'bed_files/bins_100.bed',
       sep ='\t', col.names = F, scipen = 1000)
```


## CNVRs


```{r compute CNVRs, eval = F}
dt <- cnvrs_iou(cnvs, QCtreeCNV::hg19_chr_arms, min_iou = 0.50,
                max_force_merge_rounds = 5, force_merge_min_overlap = 0.75)

saveRDS(dt, 'cnvrs.rds')
fwrite(dt[[1]], 'cnvs_with_cnvrs.txt', sep = '\t')
fwrite(dt[[2]], 'cnvrs.txt', sep = '\t')
```


```{r cnvrs BED, eval = F}
cnvs_r <- fread('cnvs_with_cnvrs.txt')
cnvrs <- fread('cnvrs.txt')
cnvrs[, name := paste0(CNVR, '_N', n)]

fwrite(cnvs_r[, .(chr, start, end, CNVR)],
       'bed_files/cnvs_with_cnvr.bed', sep ='\t', col.names = F)

fwrite(cnvrs[n >= 10, .(chr, start, end, name)],
       'bed_files/cnvrs_min10_freq.bed', sep ='\t', col.names = F)
```



# What about genes?

In the genome wide context it's always interesting to consider genes.
In this section we'll see how to access the ensembl database from R
and how to use genes as bins.

## Access Ensembl using biomaRt

```{r}

```


## Genomic enrichment analysis

```{r}

```


